Synthesizing Quotient Lenses

### 2018-03-11 papers/QuotientLenses/main.tex
finished pass over background/QREs by example

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
  @Book {Lovelace,
  Author = "Ada Lovelace",
  Title = {Notes}
}
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
  %0 Book
  %F Lovelace
  %A Ada Lovelace
  %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the ARE \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Lovelace, Ada''
and that the \codefont{space\_name} format describes ``Ada Lovelace''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. \ksf{Is all this code
correct?  Does it run through the system correctly?  I'm suspicious of
the comma\_to\_space definition which seems to be recurisive.  If it's
correct, we should comment on it.}}
\label{fig:example-qre}
\end{figure}

\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we already have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs, as well as
the language of QREs for defining equivalence relations on regular expressions.
An obvious next step in simplifying the quotient lens programming is thus to use
the bijective string lens synthesis procedure as a subroutine for a quotient lens
synthesis procedure. This new procedure should take two QREs and a set
of example input-output pairs as input, compute the canonical data formats on
both sides of the lens from the pair of QREs, map the example input-output pairs
to their canonical representatives to get a set of examples in canonical form,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the new set of examples that are in canonical form.

Indeed this basic idea is what motivates our definition of {\em QRE Lenses}.
Intuitively, just like Boomerang's quotient lenses, our QRE lenses are bijective
lenses with ``canonizers at the edges''---Figure~\ref{fig:attheedges} presents
the architecture of each lens.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{QRE Lens with ``Canonizers at the Edges''}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the get direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a unique representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the $\get$ function of a
bijective string lens $\ell_1$ to this representative. In the backward direction, $q_1$
operates similarly, but using the view QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $Q_1$ and $Q_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective string lenses is
directed by these formats, $Q_1$ and $Q_2$ are all that is required to
synthesize QRE lenses end-to-end. Indeed, we structured QRE lenses as bijective
lenses with canonizers at the edges, precisely to facilitate the use of this
synthesis algorithm.

On the other hand, our restriction that canonizers appear only at the edges
raises a key technical question:  Are we limiting the expressiveness of our
transformations by demanding all programs fit into this normal form?\bcp{I wish
we could have previewed this issue a bit more in the introduction, since it is
our main contribution.}  It turns out that we are not: any lens that uses
canonizers internally can be transformed into a lens that uses canonizers at
the edges. The main technical contribution of this paper is a proof of this
fact.

This technical investigation justifies replacing manually constructed lenses
using the QRE lens combinators with synthesized QRE lenses with canonizers at
the edges. For instance, after defining the \bibtex{} and EndNote QREs and
storing them in the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:

\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent Here, the generated  quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to intermix QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.

>>>>

\section{Background: Bijective String Languages}
\label{sec:background}
\ksf{We are inconsistent in how we typeset get/put.  Sometimes we use
codefont, sometimes we use italics of various flavors, and sometimes
we don't do anything distinctive.  It would be good to be consistent!}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
  @Book {Lovelace,
  Author = "Ada Lovelace",
  Title = {Notes}
}
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
  %0 Book
  %F Lovelace
  %A Ada Lovelace
  %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

\subsection{Specifying \bibtex{} Using QREs}
\label{subsec:qre-expressions}
\ksf{Are these QRE combinators or QRE lens combinators?  
We are currently mushy about this distinction.  
The section needs a pass to clarify this point.
The headers of this section and the next imply that this section is
about QREs and the next section is about QRE lenses, but this section
is where we introduce the QRE lens combinators, right?  If so, we need
to do something to make the structure clearer.}

@@

Sometimes there are multiple disjoint representations of the same data.
In such situations, the QRE \kw{squash} combinator creates a QRE that
allows external data to be in either format, and converts any
data in the first format to the second.
For instance, assume that
the \codefont{comma\_name} format describes ``Lovelace, Ada''
and that the \codefont{space\_name} format describes ``Ada Lovelace''
and \codefont{c\_to\_s} is a function from the first to the second.  In this
case, the following instance of \kw{squash} creates the desired canonizer.

@@

(* allow any permutation of fields interspersed with arbitrary whitespace *)
let bib_fields = perm (label, bib_author, bib_title) with (project ("," . wsp+) $\to$ ",\n")
let bibtex = "@book{" . bib_fields . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. \ksf{Is all this code
correct?  Does it run through the system correctly?  I'm suspicious of
the comma\_to\_space definition which seems to be recurisive.  If it's
correct, we should comment on it.}\ksf{In what sense is this a QRE
spec vs a QRE lens spec?}}
\label{fig:example-qre}
\end{figure}

\subsection{QRE Lenses and QRE Lens Synthesis}
\label{sec:examplesynth}
At this point we have a tool for synthesizing bijective string lenses from a
pair of regular expressions and a set of example input-output pairs
({\em Optician}), 
and we have a way of defining regular expressions with equivalence
relations indicating the essential information (QREs).
A tantalyzing possibility would be to use the bijective string lens
synthesis procedure as a subroutine for a quotient lens synthesis procedure. 
This new synthesizer would take as input source and target QREs and 
example input-output pairs, 
compute the canonical source/target formats from the QREs, 
map the example input-output pairs to their canonical representations,
and then invoke the bijective string lens synthesis procedure on the canonical data
formats and the canonical examples.

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{QRE Lens with ``Canonizers at the Edges''}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the $\get$ function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

Because the QREs $Q_1$ and $Q_2$ determine the internal formats for data after
canonization, and because the algorithm for synthesizing bijective string lenses is
directed by these formats, $Q_1$ and $Q_2$ are all that is required to
synthesize QRE lenses end-to-end.

However, our restriction that canonizers appear only at the edges
raises a key technical question: Are we limiting the expressiveness of
our transformations by demanding all programs fit into this normal
form?  It turns out that we are not: any lens that uses canonizers
internally can be transformed into a lens that uses canonizers only at
the edges. The main technical contribution of this paper
(Theorem~\ref{normal form})
is a proof of this fact.
This technical result justifies 
using synthesis to produce QRE lenses instead of manually writing
them,
which can lead to substantial savings in program complexity.
For instance, after defining the \bibtex{} and EndNote QREs and
binding them to the variables \cd{bibtex} and \cd{endnote} respectively, then
all the code in Figure~\ref{fig:example-lens} may be replaced by a single call
to the synthesis prodedure:
\ksf{What is the notation ``lens in'' below?  We should explain it
intuitively here.}
\begin{lstlisting}
let bib_to_end : (lens in bibtex $\Leftrightarrow$ endnote) =
synth bibtex $\Leftrightarrow$ endnote using {(bib_example, end_example)}
\end{lstlisting}
\noindent 
Here, the generated quotient lens synchronizes \cd{bibtex} and
\cd{endnote} formats, using \cd{bib_example} and \cd{end_example} (the two
concrete example strings given at the beginning of this section) to
disambiguate. In addition, and as we saw earlier with the definition of
\cd{c_to_s}, the synthesis procedure itself can be used to create lenses that
are in turn used to define other QREs.  The ability to interleave QRE
specification with QRE lens synthesis yields a powerful and flexible way of
creating bidirectional transformations.
